An interdisciplinary examination of stress and injury occurrence in athletes

### 2020-07-20 full_paper_v2.Rmd
pre results edits

After the eight testing sites had been identified, the testing end of the MyotonPRO (diameter = 3 mm) was positioned perpendicular to the skin on the testing site. 
A constant pre-load of 0.18 N was applied for initial compression of subcutaneous tissues. The device was programmed to deliver five consecutive impulses, separated by a one second interval [@Morgan2018]. 
For each impulse, the device computed stiffness values, with the median of the five values being saved by the device for further analysis. 
In accordance with Myoton.com, a set of five measurements with a coefficient of variation (CV) of less than 3% was accepted. 
Sets of measurements above 3 % were measured again to ensure the reliability of the data. 
The CV was calculated in real time by the device after each set of measurements. 
Measurements saved on the device were uploaded to a computer using MyotonPRO software and imported in R [@RCoreTeam2019] for further analysis. 
For each participant, the sum of all eight testing sites was calculated to provide a total lower extremity stiffness score and was used for further analysis.

@@

Instructions for the mBESS were read to each participant and a demonstration of the positions was provided by the research assistant.
For each position, participants were instructed to close their eyes, rest their hands on their iliac crests and remain as still as possible for 20-seconds. 
Participants were instructed to return to the testing position as quickly as possible if they lost their balance.
The video recording was started prior to the first stance position and stopped after all stances had been completed.
Each completed mBESS protocol took approximately 4 minutes.
Only one trial was performed to avoid familiarisation effects across the repeated measurement [cf. @Valovich2003]. 
The video recordings for each participant were imported from the recording equipment (Sony DSC-RX10) and the lead researcher scored each trial using the error identification criteria.

@@

BNs have recently been used in the sport psychology literature [@Olmedilla2018; @Fuster-Parra2017] and offer several benefits over traditional statistical analysis. 
For example, predictions can be made about any variable in the network, rather than there being a distinction between dependent and independent variables in the data, such as in linear regression models that are often used within the sport psychology literature [@Olmedilla2018; @Bittencourt2016].
Furthermore, the structure of a network can be obtained from both empirical data _and_ prior knowledge about the area of study; the latter being particularly useful when there are a large number of variables in the network, or only a small number of observations are available in the data [@Xiao-xuan2007].
In such instances, a purely data driven approach to learning the network would be time-consuming due to the large parameter space, and inefficient at identifying an approximation of the true network structure. 
Prior knowledge about dependencies between variables can therefore be included in the network structure, while still allowing a data driven approach for unknown dependencies, to improve the overall computation of the network structure [@Heckerman1995; @Xu2015].
The following sections detail the steps taken in the current study to firstly prepare the data for each network, and then obtain the structure of each network that was used for further inference.

@@

To obtain the network structure, several steps were taken to ensure that both a theoretically realistic network, and a network that was an appropriate fit to the collected data, was used for inference. 
Prior knowledge about the network structure was included by providing a list of arcs that are always _restricted_ from being in the network (blacklist), and a list of arcs that are always _included_ in the network (whitelist). 
Additionally, there are several scoring functions such as Bayesian Information Criteria (BIC) and Bayesian Dirichlet equivalent uniform (BDeu) that can be used to compare network structures with certain nodes and arcs included or excluded [@Scutari2014]. 
To account for the repeated measures design and to maximise the use of the data, pairs of complete cases (e.g., participants who completed T1 + T2, and T2 + T3) were used in a two-time Bayesian network (2TBN) structure [@Murphy2002].
In the 2TBN, variables measured T2 could depend on variables measured at T1 (e.g., T1 $\rightarrow$ T2) and variables measured at T3 could depend on variables measured at T2 (e.g., T2 $\rightarrow$ T3).
However, arcs were blacklisted between T2 $\rightarrow$ T1 and T3 $\rightarrow$ T2 to preserve the order in which data was collected. 
Variables were separated into two groups; "explanatory", for variables that did not change during the study (e.g., gender), or "independent", for variables that were measured at each time point and could vary during the study.
Independent variable names were suffixed with _1 for time point T, and _2 for time point T+1 (e.g., T1_1 $\rightarrow$ T2_2 and T2_1 $\rightarrow$ T3_2).
Formatting the data in this way meant participants who completed T1 and T2, but did not complete T3, could still be included in the analysis. 
Table \ref{tab:table2} provides an example of the formatted data. 
Participants 1 and 3 have complete data, and therefore have two rows of data each representing variables from T1 $\rightarrow$ T2 and T2 $\rightarrow$ T3, respectively. 
Participant 2 did not complete the final data collection at T3 and therefore only has one row of data representing the variables collected at T1 and T2.
In addition to the blacklisted arcs between T2 $\rightarrow$ T1 and T3 $\rightarrow$ T2, the direction of arcs was restricted between independent variables and explanatory variables (e.g., independent $\rightarrow$ explanatory); however, arcs were not restricted between explanatory $\rightarrow$ independent variables.
Finally, arc direction was restricted between specific nodes within the explanatory variables. 
Arcs from clevel $\rightarrow$ gender, nlebase $\rightarrow$ gender and nlebase $\rightarrow$ ind_team were included in the blacklist, as arcs in these directions did not make logical sense. 
All subsequent models used the same blacklist.

```{r table2, results='asis'}
# exampledata <- tibble(
#   Participant = c("1", "1", "2", "3", "3"),
#   "X_1" = c("T1", "T2", "T1", "T1", "T2"),
#   "link" = "->",
#   "X_2" = c("T2", "T3", "T2", "T2", "T3")
# )
# 
# kable(
#   exampledata,
#   caption = "Example of the data arrangement used for the network.",
#   escape = FALSE
# ) %>%
#   kable_styling(latex_options = "HOLD_position") %>%
#   row_spec(0, bold = TRUE)
```

@@

Finally, both total score and asymmetry were included in the initial network. However, visual inspection of the network revealed no arcs between bal_asym_1 or bal_asym_2 and any other node in the network.
Therefore, balance asymmetry was removed from the final network structure.
To summarise, Table \ref{tab:table3} includes the variables that were included in the final network structure.

```{r finalvars, results='asis'}
library(knitr)
library(kableExtra)
varcutoffs %>% 
  filter(!var %in% c("sdnn", "RR", "I", "GDP", "tlelg", "tlelg_1", "tlelg_2", "tlelg_3", "bal_asym")
) %>% 
  mutate(var = ifelse(var == "rmssd", "HRV", var)) %>% 
  kable("latex",
       booktabs = T,
       align = c("llcc"),
       caption = "Variables included in the final Bayesian network structure.",
       col.names = c("Variable", "Definition", "State 1", "State 2")) %>%
  kable_styling(latex_options = "HOLD_position") %>% 
  column_spec(2, width = c("5.2cm")) 
```

Preliminary network structures also revealed strong dependencies between the same variables at subsequent time points. 
For example, the probability that stiffness_1 and stiffness_2 were both "High", or both "Low" was approximately 80%. 
Including the arcs between the same variables from X_1 $\rightarrow$ X_2 did not provide any theoretically meaningful information to the network structure as the majority of participants would be either be in a "Low" or "High" state for each pair of variables in the network. Therefore, these arcs were blacked listed from the network.

To obtain the final network, the appropriate blacklist and whitelists were provided and a Tabu search algorithm identified the remaining structure of the network. 
The final network structure was obtained by averaging 1000 bootstrapped models [@Efron1993] to reduce the impact of locally optimal, but globally suboptimal network learning, and to obtain a more robust model [@Olmedilla2018].
Arcs that were present in at least 30% of the models were included in the averaged model.
The strength of each arc was determined by the percentage of models that the arc was included in, independent of the arc's direction.
An arc strength of 1 indicates that the arc is always present in the network, with the value decreasing as arcs are found in fewer networks. 
In the respective study arcs above 0.5 were considered "significant" with arcs below 0.5 and above 0.3 "non-significant" [@Scutari2013]. 
Arcs below 0.3 were not included in the model.
The full table of arc strengths for the first and second network are available in S2 Table and S3 Table respectively.

@@

Conditional probability queries (CPQ) were used to perform inference on both network structures.
To conduct a CPQ, the joint probability distribution of the nodes was modified to include a new piece of evidence.
The query allows the odds of a particular node state (e.g., injured_1 = "injured") to be calculated based on the new evidence. 
CPQs were performed using a likelihood weighting approach; a form of importance sampling where random observations are generated from the probability distribution in such a way that all observations match the evidence given in the query. 
The algorithm then re-weights each observation based on the evidence when computing the conditional probability for the query [@Scutari2014].
Inference was first performed on arcs that had a strength greater than 0.50 between the explanatory variables and independent variables and between different independent variables in the network. 
Of particular interest in the current study were the variables that were connected to "injured" nodes.
To examine the variables that were associated with injured nodes in the network, the Markov blanket of "injured_1" and "injured_2" were examined. 
A Markov blanket contains all the nodes that make the node of interest conditionally independent from the rest of the network [@Fuster-Parra2017].
CPQ's were used to determine what effect the variables in the Markov blanket of injured nodes had on the probability of the injured node being in the "injured" state.

@@

For the second network, continuous variables suffixed with _1 were subtracted from variables suffixed with _2 to obtain the difference between variables measured at time points T1 $\rightarrow$ T2 and T2 $\rightarrow$ T3. 
Independent variables were then standardised (mean-centred at 0) to allow relative changes between variables to be compared. 
The "injured" variable was also modified to represent whether a participant had sustained an injury at any point over the duration of the study or were healthy for the duration of the study.
Setting the data up in this way enabled the construction of a network that explicitly modelled the _amount_ of change within variables between time points, as opposed to the first network that would only have captured changes when the median threshold was crossed from "Low" to "High".

@@

<!-- The conditional probabilities based on all the variables in injured_2 Markov blanket are presented in Table \ref{tab:table9}. -->
<!-- Again, only the three lowest and highest probabilities are shown (complete results in S5 Table). -->
<!-- The combination "Low" FFFS_1, "High" stiffness_2, "High" balance resulted in the greatest  probability of injured_2 being "injured", with the highest probability of injury being `r max(query4$prob)`. -->
<!-- With all other variables held in the "High" state, the probability of injured_2 being "injured" rose from 0.14 to 0.34 when FFFS_1 was in the "Low" compared to "High" state. -->
<!-- The combination of "Low" stiffness, "Low" balance and "High" FFFS resulted in the lowest probability of injured_2 being "injured". -->

<!-- ```{r table9, results="asis" } -->

<!-- query4 %>% -->
<!--   select(7,2:6) %>% -->
<!--   rename(Probability = "prob") %>% -->
<!--   `colnames<-`(str_replace_all(colnames(.), "_", "\\\\_")) %>% -->
<!--   mutate_all(~str_replace_all(., "_", "\\\\_"))%>% -->
<!--   kable( -->
<!--     ., -->
<!--     caption = "Highest and lowest probability of injured\\_2 being in the 'injured' state, conditional on the all variables in the Markov blanket for injured\\_2.", -->
<!--     escape = F -->
<!--   ) %>% -->
<!--   kable_styling(latex_options = "HOLD_position") %>% -->
<!--   pack_rows("Highest", 1, 3) %>% -->
<!--   pack_rows("Lowest", 4, 6) %>% -->
<!--   row_spec(0,bold=TRUE) -->
<!-- ``` -->

@@

>>>>

After the eight testing sites had been identified, the testing end of the MyotonPRO (diameter = 3 mm) was positioned perpendicular to the skin on the testing site. 
A constant pre-load of 0.18 N was applied for initial compression of subcutaneous tissues.
The device was programmed to deliver five consecutive impulses, separated by a one-second interval [@Morgan2018]. 
For each impulse, the device computed passive stiffness values, with the median of the five values being saved by the device for further analysis. In accordance with Myoton.com, a set of five measurements with a coefficient of variation (CV) of less than 3% was accepted. Sets of measurements above 3% were measured again to ensure data reliability. 
Measurements were uploaded using MyotonPRO software and imported in R [@RCoreTeam2019] for further analysis. For each participant, the sum of all eight testing sites was calculated to provide a total lower extremity stiffness score and was used for further analysis

@@

Instructions for the mBESS were read to each participant and a demonstration of the positions was provided by the research assistant.
For each position, participants were instructed to close their eyes, rest their hands on their iliac crests and remain as still as possible for 20-seconds. 
Participants were instructed to return to the testing position as quickly as possible if they lost their balance.
The video recording was started prior to the first stance position and stopped after all stances had been completed.
Each completed mBESS protocol took approximately four-minutes.
Only one trial was performed to avoid familiarisation effects across the repeated measurement [cf. @Valovich2003]. 
The video recordings for each participant were imported from the recording equipment (Sony DSC-RX10) and the lead researcher scored each trial using the error identification criteria.

@@

BNs have recently been used in the sport psychology literature [@Olmedilla2018; @Fuster-Parra2017] and offer several benefits over traditional statistical analysis. 
For example, predictions can be made about any variable in the network, rather than there being a distinction between dependent and independent variables in the data, such as in linear regression models that are often used within the sport psychology literature [@Olmedilla2018; @Bittencourt2016].
Furthermore, the structure of a network can be obtained from both empirical data _and_ prior knowledge about the area of study; the latter being particularly useful when there are a large number of variables in the network, or only a small number of observations are available in the data [@Xiao-xuan2007].
In such instances, a purely data driven approach to learning the network would be time-consuming due to the large parameter space, and inefficiency at identifying an approximation of the true network structure. 
Prior knowledge about dependencies between variables can therefore be included in the network structure, while still allowing a data driven approach for unknown dependencies, to improve the overall computation of the network structure [@Heckerman1995; @Xu2015].
The following sections detail the steps taken in the current study to firstly prepare the data for each network, and then obtain the structure of each network that was used for further inference.

@@

To obtain the network structure, several steps were taken to ensure that both a theoretically realistic network, and a network that was an appropriate fit to the collected data, was used for inference. 
Prior knowledge about the network structure was included by providing a list of arcs that are always _restricted_ from being in the network (blacklist), and a list of arcs that are always _included_ in the network (whitelist). 
Additionally, there are several scoring functions such as Bayesian Information Criteria (BIC) and Bayesian Dirichlet equivalent uniform (BDeu) that can be used to compare network structures with certain nodes and arcs included or excluded [@Scutari2014]. 
To account for the repeated measures design employed in this study and to maximise the use of the data, pairs of complete cases (e.g., participants who completed T1 + T2, and T2 + T3) were used in a two-time Bayesian network (2TBN) structure [@Murphy2002].
In the 2TBN, variables measured T2 could depend on variables measured at T1 (e.g., T1 $\rightarrow$ T2) and variables measured at T3 could depend on variables measured at T2 (e.g., T2 $\rightarrow$ T3).
However, arcs were blacklisted between T2 $\rightarrow$ T1 and T3 $\rightarrow$ T2 to preserve the order in which data was collected. 
Variables were separated into two groups; "explanatory", for variables that did not change during the study (e.g., gender), or "independent", for variables that were measured at each time point and could vary during the study.
Independent variable names were suffixed with _1 for time point T, and _2 for time point T+1 (e.g., T1_1 $\rightarrow$ T2_2 and T2_1 $\rightarrow$ T3_2).
Formatting the data in this way meant participants who completed T1 and T2, but did not complete T3, could still be included in the analysis. 
Table \ref{tab:table2} provides an example of the formatted data and demonstrates that participants 1 and 3 have complete data, and therefore have two rows of data each representing variables from T1 $\rightarrow$ T2 and T2 $\rightarrow$ T3, respectively. 
Participant 2 did not complete the final data collection at T3 and therefore only has one row of data representing the variables collected at T1 and T2.
In addition to the blacklisted arcs between T2 $\rightarrow$ T1 and T3 $\rightarrow$ T2, the direction of arcs was restricted between independent variables and explanatory variables (e.g., independent $\rightarrow$ explanatory); however, arcs were not restricted between explanatory $\rightarrow$ independent variables.
Finally, arc direction was restricted between specific nodes within the explanatory variables. 
Arcs from clevel $\rightarrow$ gender, nlebase $\rightarrow$ gender and nlebase $\rightarrow$ ind_team were included in the blacklist, as arcs in these directions did not make logical sense. 
All subsequent models used the same blacklist.

@@

Finally, both total score and asymmetry for balance were included in the initial network. However, visual inspection of the network revealed no arcs between bal_asym_1 or bal_asym_2 and any other node in the network.
Therefore, balance asymmetry was removed from the final network structure.
To summarise, Table \ref{tab:table3} reports the variables that were included in the final network structure.

Preliminary network structures also revealed strong dependencies between the same variables at sequential time points. For example, the probability that stiffness_1 and stiffness_2 were both “High”, or both “Low” was approximately 80%. Including the arcs between the same variables from X_1 → X_2 did not provide any theoretically meaningful information to the network structure as the majority of participants would either be in a “Low” or “High” state for each pair of variables in the network. Therefore, these arcs were blacklisted from the network.
To obtain the final network, the appropriate blacklist and whitelists were provided and a Tabu search algorithm identified the remaining structure of the network. 
The final network structure was obtained by averaging 1000 bootstrapped models [@Efron1993] to reduce the impact of locally optimal, but globally suboptimal network learning, and to obtain a more robust model [@Olmedilla2018]. Arcs that were present in at least 30% of the models were included in the averaged model. 
The strength of each arc was determined by the percentage of models that the arc was included in, independent of the arc’s direction. An arc strength of 1 indicated that the arc was always present in the network, with the value decreasing as arcs were found in fewer networks. In the respective study arcs above 0.5 were considered “significant” with arcs below 0.5 and above 0.3 “non-significant” [@Scutari2013]. 
Arcs below 0.3 were not included in the model. The full table of arc strengths for the first and second network are available in S2 Table and S3 Table respectively.

@@

Conditional probability queries (CPQ) were used to perform inference on both network structures. To conduct a CPQ, the joint probability distribution of the nodes was modified to include a new piece of evidence. The query allows the odds of a particular node state (e.g., injured_1 = “injured”) to be calculated based on the new evidence. CPQs were performed using a likelihood weighting approach; a form of importance sampling where random observations are generated from the probability distribution in such a way that all observations match the evidence given in the query. The algorithm then re-weights each observation based on the evidence when computing the conditional probability for the query [@Scutari2014]. Inference was first performed on arcs that had a strength greater than 0.5 between the explanatory variables and independent variables and between different independent variables in the network. Of particular interest were the variables that were connected to “injured” nodes, which were examined in the network using the Markov blanket of “injured_1” and “injured_2”. A Markov blanket contains all the nodes that make the node of interest conditionally independent from the rest of the network [@Fuster-Parra2017]. CPQ’s were used to determine what effect the variables in the Markov blanket of injured nodes had on the probability of the injured node being in the “injured” state.

@@

Change scores between timepoint 1 and timepoint 2 were standardized to allow relative changes between variables to be compared.

For the second network, change scores for continuous variables between  T1 $\rightarrow$  T2 and  T2 $\rightarrow$  T3 were standardized to allow relative changes between variables to be compared.
The "injured" variable was also modified to represent whether a participant had sustained an injury at any point over the duration of the study or were healthy for the duration of the study.
Setting the data up in this way enabled the construction of a network that explicitly modelled the _amount_ of change within variables between time points, as opposed to the first network that only captured changes when the median threshold was crossed from "Low" to "High".

@@

The conditional probabilities based on all the variables in injured_2 Markov blanket are presented in Table \ref{tab:table9}.
Again, only the three lowest and highest probabilities are shown (complete results in S5 Table).
The combination "Low" FFFS_1, "High" stiffness_2, "High" balance resulted in the greatest  probability of injured_2 being "injured", with the highest probability of injury being `r max(query4$prob)`.
With all other variables held in the "High" state, the probability of injured_2 being "injured" rose from 0.14 to 0.34 when FFFS_1 was in the "Low" compared to "High" state.
The combination of "Low" stiffness, "Low" balance and "High" FFFS resulted in the lowest probability of injured_2 being "injured".

@@

# Tables

```{r table2, results='asis'}
exampledata <- tibble(
  Participant = c("1", "1", "2", "3", "3"),
  "X\\_1" = c("T1", "T2", "T1", "T1", "T2"),
  "X\\_2" = c("T2", "T3", "T2", "T2", "T3")
)

kable(
  exampledata,
  caption = "Example of the data arrangement used for the network.",
  escape = FALSE
) %>%
  kable_styling(latex_options = "HOLD_position") %>%
  row_spec(0, bold = TRUE)
```

```{r table3, results='asis'}
varcutoffs %>% 
  filter(!var %in% c("sdnn", "RR", "I", "GDP", "tlelg", "tlelg_1", "tlelg_2", "tlelg_3", "bal_asym")
) %>% 
  mutate(var = ifelse(var == "rmssd", "HRV", var)) %>% 
  kable("latex",
       align = c("llcc"),
       caption = "Variables included in the final Bayesian network structure.",
       col.names = c("Variable", "Definition", "State 1", "State 2")) %>%
  kable_styling(latex_options = "HOLD_position") %>% 
  column_spec(2, width = c("5.2cm")) 
```

```{r table9, results="asis" }

query4 %>%
  select(7,2:6) %>%
  rename(Probability = "prob") %>%
  `colnames<-`(str_replace_all(colnames(.), "_", "\\\\_")) %>%
  mutate_all(~str_replace_all(., "_", "\\\\_"))%>%
  kable(
    .,
    caption = "Highest and lowest probability of injured\\_2 being in the 'injured' state, conditional on the all variables in the Markov blanket for injured\\_2.",
    escape = F
  ) %>%
  kable_styling(latex_options = "HOLD_position") %>%
  pack_rows("Highest", 1, 3) %>%
  pack_rows("Lowest", 4, 6) %>%
  row_spec(0,bold=TRUE)
```
